---
title: "Pseudo-bulk based comparison of cell types in two Seurat objects"
author: "Minzhe Guo, Cheng Jiang"
date: 'Compiled: `r Sys.Date()`'
output: 
  html_document:
  theme: united
  df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  time_it = TRUE
)
```

### Introduction
In this document, We will be using two publicly-available human PBMC datasets as an example to demonstrate the comparison of cell types in two Seurat objects based on pseudo-bulk expression profiles and  highly variable genes.

### Loading pacakge
First, load in Seurat and packages that will be used in this demo. 
```{r Libraries}
library(Seurat)
library(pheatmap)
library(ggplot2)
library(patchwork)
```

### Data preparation
The example datasets that we will use for demonstration are available through SeuratData package (https://github.com/satijalab/seurat-data). 
```{r Read Data}
# load the SeuratData pacakge
library(SeuratData)
```

Let's download the two example human PBMC datasets from SeuratData using InstallData(). 
````{r download data}
InstallData("ifnb")
InstallData("pbmcsca")
```

Load the first dataset "ifnb" and assign it to the variable "data1". 
Check the object and avaialble cell metadata information. 
We can see that the data is already a Seurat object, containing an RNA assay of 14,503 features in 13,999 cells.
The celltype annotations are in the  "seurat_annotations" column of meta.data.
The data is likely not normalized, as the max value of the "data" slot is a large integer. 
Run NormalizedData() to normalize the data.

```{r}
data("ifnb")
data1 = ifnb

data1

head(data1@meta.data)

max(data1@assays$RNA@data)

data1 = NormalizeData(data1, verbose = F)
```

Load the second dataset "pbmcsca" and assign it to the variable "data2". 
Check the object and avaialble cell metadata information. 
We can see that the data is also a Seurat object, containing an RNA assay of 33,694 features in 31,021 cells. 
The celltype annotations are in the  "CellType" column of meta.data.
The data is also likely not normalized, as the max value of the "data" slot is a large integer. 
Run NormalizedData() to normalize the data.
```{r}
data("pbmcsca")
data2 = pbmcsca

data2

head(data2@meta.data)

max(data2@assays$RNA@data)

data2 = NormalizeData(data2, verbose = F)
```

### Create pseudo-bulk profiles

Create a pseudo-bulk profile for each cell type in each dataset. 
A pseudo-bulk profile is comprised of each gene's averaged expression in a cell type.

```{r}
# set the cell type annotation as the active identity of the Seurat object
data1 = SetIdent(data1, value=data1@meta.data$seurat_annotations)
data2 = SetIdent(data2, value=data2@meta.data$CellType)

# Use AverageExpression() to calculate each gene's average expression in each cell type.
# By setting return.seurat=T, the results will be returned as a Seurat object
data1.avg = AverageExpression(data1, assay="RNA", slot="data", return.seurat = T)
data2.avg = AverageExpression(data2, assay="RNA", slot="data", return.seurat = T)
```
### Find highly variable genes
Find highly variable genes that have expression in both datasets for the correlation calculation.

```{r}
# first, find top 2000 most highly variable genes (HVGs) within each dataset
data1.avg = FindVariableFeatures(data1.avg, nfeatures = 2000)
data2.avg = FindVariableFeatures(data2.avg, nfeatures = 2000)

# union the HVGs from the two datasets
hvg.use = union(data1.avg@assays$RNA@var.features, 
                   data2.avg@assays$RNA@var.features)
cat(length(hvg.use), "HVGs identitied from the two datasets\n")

# keep HVGs that have expression in both datasets
hvg.use = hvg.use[which(hvg.use %in% rownames(data1.avg@assays$RNA@data))]
hvg.use = hvg.use[which(hvg.use %in% rownames(data2.avg@assays$RNA@data))]

cat("Use", length(hvg.use), "HVGs that have expression in both datasets\n")

```

Scale pseudo-bulk expression of the HVGs
```{r}
data1.avg = ScaleData(data1.avg, features = hvg.use)
data2.avg = ScaleData(data2.avg, features = hvg.use)
```

Create a matrix that combines the scaled pseudo-bulk expression of HVGs in different datasets 
```{r}
# get the scaled pseudo-bulk expression of the HVGs
data1.data = data1.avg@assays$RNA@scale.data
data2.data = data2.avg@assays$RNA@scale.data

# add suffix to column names (cell types) to distinguish the data from different datasets
colnames(data1.data) = paste0(colnames(data1.data), ".data1")
colnames(data2.data) = paste0(colnames(data2.data), ".data2")

# combine the data to create an expression matrix
expr_mat = cbind(data1.data, data2.data[rownames(data1.data), ])

# make sure no missing values generated
any(is.na(expr_mat))
```

### Correlations and Visualization
```{r Correlation, fig.width=10, fig.height=11}

# calcualte the correlations of cell types using scaled pseudo-bulk expression of HVGs
cor_mat = cor(expr_mat)

# visualize the correlations using pheatmap;
# by setting display_numbers = T, the correlation values will be shown on the heatmap;
# use the round() function to round the correlation values to two decimal places for better display
g = pheatmap::pheatmap(round(cor_mat, digits = 2), 
                       clustering_method = "complete", 
                       display_numbers = T
                       )


# save the heatmap to a tif file
tiff(filename = "cor_heatmap.tif", width=12, height=11, res=300, units="in", compression="lzw")
print(g)
dev.off()

# save the HVG expression matrix and the correlation matrix as source data
sourcedata = list(hvg.expression=expr_mat, correlations=cor_mat)
save(sourcedata, file="sourcedata.rda")

```


<details open>

<summary>

**Session Info**

</summary>

```{r}
sessionInfo()
```